Supplementary material for "Hack Weeks as a model for Data Science Education and Collaboration"

### 2018-05-12 intro.tex
Added N to figures, also updated some text

This looser structure is designed to facilitate a multi-directional transfer of knowledge between a diverse group of people in different stages of their career and with different backgrounds (see also \ref{sec:participants}). The flexibility of our approach results in shifts along the continuum of pedagogy and open project work across hackweek disciplines, and from year to year within a specific discipline (Figure \ref{fig:hackspectrum}).

>>>>

This looser structure is designed to facilitate a multi-directional transfer of knowledge between a diverse group of people in different stages of their career and with different backgrounds (see also section Participants below). The flexibility of our approach results in shifts along the continuum of pedagogy and open project work across hackweek disciplines, and from year to year within a specific discipline (Figure \ref{fig:hackspectrum}).

### 2018-05-12 participants.tex
Added N to figures, also updated some text

Hack weeks differ from traditional conferences or summer schools in that knowledge transfer occurs across many levels of seniority, disciplinary boundaries, and novelty of the topics discussed.
In addition, a substantial amount of hack week content is generated during the event itself, requiring active participation from attendees.
In our experience, maximizing learning outcomes and collaborative exchanges at hack weeks requires a participant group that is diverse across categories of minority status, geographical origin, gender, discipline and career track.

Traditional selection processes that rely heavily on internal heuristics of reviewers, especially thost that consider characteristics peripheral to the evaluation criteria, are often fraught personal and structural biases~\cite[e.g.][]{sunstein2015wiser}. 
In order to maximize diversity and minimize bias, we advocate for a selection process that is as quantitative and transparent as possible~\cite{sunstein2015wiser}, enabling participants to hold organizers accountable for their selection decisions.
This requires laying out a definition of successful participation, what criteria must be met in order to maximize the likelihood of success, and how those criteria will be assessed given the information about the candidates selected during the application stage.

@@

If merit-based selection is part of the evaluation process, organizers face the crucial decision whether to assess merit blindly or not. Because human decision makers tend to be swayed by unrelated characteristics including name~\cite{bertrand2004} or gender~\cite{mossracusin2012}, an initial merit selection blinded to demographic characteristics can be an effective way to counteract certain biases related to internal heuristics. A merit selection could then be performed via scores given independently by members of the organizing committee based on a set of pre-defined, explicit selection criteria. This type of blinded procedure tends to reduce biases during participant selection when committees would otherwise not consider diversity during their selection [REF]. However, a blind selection purely on a figure of merit will likely still be biased if some groups of applicants have been systematically disadvantaged in certain ways: for example, if workshop organizers require a minimum level of programming experience, this could disadvantage candidates who have had fewer opportunities to learn programming due to structural inequalities.

Blinding to demographics can be counterproductive if it excludes participants who might have had less exposure to certain technologies or fewer opportunities to learn certain skills. Additionally, blinding has been found to have negative effects on diversity for committees that have already had a strong commitment to diversity, because they often correct for structural inequalities by considering demographic variables during merit selection~\cite{behaghel2015unintended}.
In this case, it may be beneficial to construct selection criteria that explicitly consider diversity and inclusivity (as Neuro Hack Week has done; see also SM, Section 2.1). 
Because systemic biases likely also enter at the application stage (where underrepresented groups may be less likely to apply) organizers should consider oversampling traditionally disenfranchised groups compared to the population of applicants.

>>>>

Hack weeks differ from traditional conferences or summer schools in that knowledge transfer occurs across many levels of seniority, disciplinary boundaries, and novelty of the topics discussed.
In addition, a substantial amount of hack week content is generated during the event itself, requiring active participation from attendees.
In our experience, maximizing learning outcomes and collaborative exchanges at hack weeks requires a participant group that is diverse across categories of minority status, geographical origin, gender, discipline and career track, among others.

Traditional selection processes that rely heavily on internal heuristics of reviewers, especially those that consider characteristics peripheral to the evaluation criteria, are often fraught personal and structural biases~\cite[e.g.][]{sunstein2015wiser}. 
In order to maximize diversity and minimize bias, we advocate for a selection process that is as quantitative and transparent as possible~\cite{sunstein2015wiser}, enabling participants to hold organizers accountable for their selection decisions.
This requires laying out a definition of successful participation, what criteria must be met in order to maximize the likelihood of success, and how those criteria will be assessed given the information about the candidates selected during the application stage.

@@

If merit-based selection is part of the evaluation process, organizers face the crucial decision whether to assess merit blindly or not. Because human decision makers tend to be swayed by unrelated characteristics including name~\cite{bertrand2004} or gender~\cite{mossracusin2012}, an initial merit selection blinded to demographic characteristics can be an effective way to counteract certain biases related to internal heuristics. A merit selection could then be performed via scores given independently by members of the organizing committee based on a set of pre-defined, explicit selection criteria. This type of blinded procedure tends to reduce biases during participant selection when committees would otherwise not consider diversity during their selection [REF].

However, a blind selection purely on a figure of merit will be counterproductive if it excludes participants who might have had less exposure to certain technologies or fewer opportunities to learn certain skills: for example, if workshop organizers require a minimum level of programming experience, this could disadvantage candidates who have had fewer opportunities to learn programming due to structural inequalities. 
Additionally, blinding has been found to have negative effects on diversity for committees that have already had a strong commitment to diversity, because they often correct for structural inequalities by considering demographic variables during merit selection~\cite{behaghel2015unintended}.
In this case, it may be beneficial to construct selection criteria that explicitly consider diversity and inclusivity (as Neuro Hack Week has done; see also SM, Section 2.1). 
Because systemic biases likely also enter at the application stage (where underrepresented groups may be less likely to apply) organizers should consider oversampling traditionally disenfranchised groups compared to the population of applicants.

### 2018-05-12 results.tex
Added intro changes from Anthony
Added N to figures, also updated some text

\begin{table*}
%\begin{table*}[hbtp]
%\renewcommand{\arraystretch}{1.3}
\footnotesize
\caption{}
\begin{threeparttable} 
\begin{tabularx}{18cm}{p{3.4cm}p{0.3cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}}
\toprule
\textbf{Question} & \multicolumn{6}{c}{\textbf{Astro Hack Week}} & \multicolumn{6}{c}{\textbf{GeoHackWeek}} & \multicolumn{6}{c}{\textbf{NeuroHackWeek}} \\ \midrule
 & $N$ & $\mathrm{dof}_{\chi^2}^\emph{a}$ & $\chi^2$ & $p$ & $\mathrm{dof}_{\phi_c}^\emph{b}$ & $\phi_c$ & $N$ & $\mathrm{dof}_{\chi^2}^\emph{a}$ & $\chi^2$ & $p$ & $\mathrm{dof}_{\phi_c}^\emph{b}$ & $\phi_c$ & $N$ & $\mathrm{dof}_{\chi^2}^\emph{a}$ & $\chi^2$ & $p$ & $\mathrm{dof}_{\phi_c}^\emph{b}$ & $\phi_c$ \\ \midrule
\multicolumn{16}{l}{\textbf{Comparison with career stage}}  \\ \midrule
New topics, tools, methods (a) & 61 & 10 & 5.64 & 0.85 & 1.93 & 0 & 36 & 6 & 2.51 & 0.87 & 1.88 & 0 & 41& 10 & 15.53 & 0.15 & 1.9 & 0.23 \\
Improvement of research (b) & 61 & 6 &  9.89 & 0.13 & 1.93 & 0.18 & 35 & 4 & 12.41 & 0.02 & 1.88 & 0.35 & 40 & 6 & 7.42 & 0.28 & 1.90 & 0.13 \\
Better science (c) & 59 & 6 & 3.79 & 0.70 & 1.93 & 0 & 35 & 4 & 10.61 & 0.03 & 1.88 & 0.31 & 40 & 4 &  5.06 & 0.29 & 1.90 & 0.11\\ 
Teaching (d) & 61 & 10 & 13.65 & 0.19 & 1.93 & 0.17 & 36 & 10 & 13.68 & 0.19 & 1.88 & 0.22 & 40 & 10 & 6.90 & 0.74 & 1.90 & 0 \\ \midrule
\multicolumn{16}{l}{\textbf{Gender Identity}} \\ \midrule
Improvement of research (b) & 60 & 3 & 2.08 & 0.56 & 0.98 & 0 & 35 & 2 & 5.26 & 0.07 & 0.97 & 0.30 & 40 & 3 & 7.62 & 0.05 & 0.97 & 0.34\\
Teaching (d) & 61 & 5 & 7.30 & 0.20 & 0.98 & 0.19 & 36 & 5 & 12.71 & 0.26 & 0.97 & 0.46 & 40 &5 & 3.17 & 0.67 & 0.97 & 0 \\
Connection building (e) & 61 & 4 & 3.70 & 0.45 & 0.98 & 0 & 36 & 3 & 5.89 & 0.12 & 0.97 & 0.28 & 41 & 3 & 4.42 & 0.22 & 0.98 & 0.18\\
Valued contributions (f) & 57 & 4 & 5.98 & 0.20 & 0.98 & 0.18 & 34 & 5 & 9.46 & 0.09 & 0.97 & 0.36 & 38 & 5 & 6.75 & 0.34 & 0.97 & 0.21 \\ \midrule
\multicolumn{16}{l}{\textbf{Racial/ethnic Identity}} \\ \midrule
Improvement of research (b)  & 59 & 3 & 0.97 & 0.81 & 0.98 & 0 & 34 &2 & 0.01 & 0.99 & 0.97 & 0 & 39 & 3 & 2.79 & 0.43 & 0.97 & 0 \\
Teaching (d) & 61 & 5 & 5.63 & 0.34 & 0.98 & 0.09 & 34 & 4 & 1.29 & 0.86 & 0.97 & 0 & 39 & 5 & 3.06 & 0.69 & 0.97 & 0 \\
Connection building (e) & 61 & 4 & 1.64 & 0.80 & 0.98 & 0 & 34 & 3 & 8.42 & 0.04 & 0.97 & 0.40 & 40 & 3 & 3.49 & 0.32 & 0.97 & 0.10 \\
Valued contributions (f) & 58 & 4 & 13.14 & 0.01 & 0.98 & 0.40 & 32 & 5 & 2.49 & 0.78 & 0.96 & 0 & 37 & 5 & 4.05 & 0.54 & 0.97 & 0 \\ 
\bottomrule
\end{tabularx}
   \begin{tablenotes}
      \item{Respondents to a post-event evaluation survey indicate their level of agreement on a six-point Likert-type scale to a range of statements (corresponding to Figure \ref{fig:survey}). Results are for hack weeks held in 2016 and 2017 for AHW and in 2016 for GHW and NHW, respectively. The number of respondents for each question is included in the table above. Additional figures showing the distributions of scores are available in the SM, Section 6.}
      \item[\emph{a}]{The degrees of freedom for Cram\'{e}r's V corrected according to~\citep{bergsma2013}}
      \item[\emph{b}]{While the range of Likert items was the same for every question, the degrees of freedom for $\chi^2$ and Cram\'{e}r's V may differ depending on the number of items with non-zero responses, and the number of non-zero items in the independent variable.}

@@

\begin{figure*}[h!]
%\begin{center}
%\begin{subfigure}[t]{\textwidth}
\centering
%\caption{}
\includegraphics[width=\textwidth]{f2.eps}
%\end{subfigure}
\caption{Post-workshop surveys from three hack weeks: participants in the 2016 astro-, geo- and neuro- hack weeks responded to questions assessing their experiences. For GHW and NHW, response rates were 100\%, with $N_{\mathrm{GHW}}= 83$ and $N_{\mathrm{NHW}} = 86$, respectively. At AHW, the combined response rate for both years was 76\%, or $N_{\mathrm{AHW}} = 72$ out of $94$ total attendees. We report here about results in three different domains: the development of technical skills (a -- c), collaboration and teaching (d -- f), and shifts in attitudes towards reproducibility and open science (g, h).}
\label{fig:survey}
%\end{center}
\end{figure*}

@@

Given the broad diversity in skills and backgrounds of participants admitted to our hack weeks, a dependence of learning outcomes and teaching on career stage is plausible. At the same time, we put forth the hypothesis earlier in the this paper that peer learning is a major mode of knowledge transfer during hack weeks, and that this type of learning takes place at all career stages. However, we find no strong evidence for a significant difference between early-career researchers and senior participants for any of the three hack weeks for questions regarding learning outcomes of new tools and topics (a), improvements in day-to-day research (b) or overall improvements in science (c) (all statistical results are presented in Table \ref{tab:survey}). Only for GHW we find an indication that early-career researchers agree more strongly that the hack week improved their day-to-day research (p=0.02) with a large effect size of $\phi_c = 0.4$ (following~\citep{cohen1988}). 
Similarly, we find no strong indication that late-career researchers self-report a higher level of teaching at hack weeks compared to early-career researchers (d).

>>>>

\begin{figure*}[h!]
%\begin{center}
%\begin{subfigure}[t]{\textwidth}
\centering
%\caption{}
\includegraphics[width=\textwidth]{f2.eps}
%\end{subfigure}
\caption{Post-workshop surveys from three hack weeks: participants in the 2016 astro-, geo- and neuro- hack weeks responded to questions assessing their experiences. The overall response rates for GHW and NHW were 100\%, with $N_{\mathrm{GHW}}= 83$ and $N_{\mathrm{NHW}} = 86$, respectively. At AHW, the combined overall response rate for both years was 76\%, or $N_{\mathrm{AHW}} = 72$ out of $94$ total attendees. Note that individual response rates may be slightly lower (as some questions were optional) and are reported in the panel titles for AHW, GHW and NHW, respectively. We report here about results in three different domains: the development of technical skills (a -- c), collaboration and teaching (d -- f), and shifts in attitudes towards reproducibility and open science (g, h).}
\label{fig:survey}
%\end{center}
\end{figure*}

\begin{table*}
%\begin{table*}[hbtp]
%\renewcommand{\arraystretch}{1.3}
\footnotesize
\caption{}
\begin{threeparttable} 
\begin{tabularx}{18cm}{p{3.4cm}p{0.3cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}}
\toprule
\textbf{Question} & \multicolumn{6}{c}{\textbf{Astro Hack Week}} & \multicolumn{6}{c}{\textbf{GeoHackWeek}} & \multicolumn{6}{c}{\textbf{NeuroHackWeek}} \\ \midrule
 & $N$ & $\mathrm{dof}_{\chi^2}^\emph{a}$ & $\chi^2$ & $p$ & $\mathrm{dof}_{\phi_c}^\emph{b}$ & $\phi_c$ & $N$ & $\mathrm{dof}_{\chi^2}^\emph{a}$ & $\chi^2$ & $p$ & $\mathrm{dof}_{\phi_c}^\emph{b}$ & $\phi_c$ & $N$ & $\mathrm{dof}_{\chi^2}^\emph{a}$ & $\chi^2$ & $p$ & $\mathrm{dof}_{\phi_c}^\emph{b}$ & $\phi_c$ \\ \midrule
\multicolumn{16}{l}{\textbf{Comparison with career stage (early-career stage; faculty and research scientists; other)}}  \\ \midrule
New topics, tools, methods (a) & 61 & 10 & 5.64 & 0.85 & 1.93 & 0 & 36 & 6 & 2.51 & 0.87 & 1.88 & 0 & 41& 10 & 15.53 & 0.15 & 1.9 & 0.23 \\
Improvement of research (b) & 61 & 6 &  9.89 & 0.13 & 1.93 & 0.18 & 35 & 4 & 12.41 & 0.02 & 1.88 & 0.35 & 40 & 6 & 7.42 & 0.28 & 1.90 & 0.13 \\
Better science (c) & 59 & 6 & 3.79 & 0.70 & 1.93 & 0 & 35 & 4 & 10.61 & 0.03 & 1.88 & 0.31 & 40 & 4 &  5.06 & 0.29 & 1.90 & 0.11\\ 
Teaching (d) & 61 & 10 & 13.65 & 0.19 & 1.93 & 0.17 & 36 & 10 & 13.68 & 0.19 & 1.88 & 0.22 & 40 & 10 & 6.90 & 0.74 & 1.90 & 0 \\ \midrule
\multicolumn{16}{l}{\textbf{Gender Identity (non-minority; minority$^\emph{c}$)}} \\ \midrule
Improvement of research (b) & 60 & 3 & 2.08 & 0.56 & 0.98 & 0 & 35 & 2 & 5.26 & 0.07 & 0.97 & 0.30 & 40 & 3 & 7.62 & 0.05 & 0.97 & 0.34\\
Teaching (d) & 61 & 5 & 7.30 & 0.20 & 0.98 & 0.19 & 36 & 5 & 12.71 & 0.26 & 0.97 & 0.46 & 40 &5 & 3.17 & 0.67 & 0.97 & 0 \\
Connection building (e) & 61 & 4 & 3.70 & 0.45 & 0.98 & 0 & 36 & 3 & 5.89 & 0.12 & 0.97 & 0.28 & 41 & 3 & 4.42 & 0.22 & 0.98 & 0.18\\
Valued contributions (f) & 57 & 4 & 5.98 & 0.20 & 0.98 & 0.18 & 34 & 5 & 9.46 & 0.09 & 0.97 & 0.36 & 38 & 5 & 6.75 & 0.34 & 0.97 & 0.21 \\ \midrule
\multicolumn{16}{l}{\textbf{Racial/ethnic Identity (white; non-white)}} \\ \midrule
Improvement of research (b)  & 59 & 3 & 0.97 & 0.81 & 0.98 & 0 & 34 &2 & 0.01 & 0.99 & 0.97 & 0 & 39 & 3 & 2.79 & 0.43 & 0.97 & 0 \\
Teaching (d) & 61 & 5 & 5.63 & 0.34 & 0.98 & 0.09 & 34 & 4 & 1.29 & 0.86 & 0.97 & 0 & 39 & 5 & 3.06 & 0.69 & 0.97 & 0 \\
Connection building (e) & 61 & 4 & 1.64 & 0.80 & 0.98 & 0 & 34 & 3 & 8.42 & 0.04 & 0.97 & 0.40 & 40 & 3 & 3.49 & 0.32 & 0.97 & 0.10 \\
Valued contributions (f) & 58 & 4 & 13.14 & 0.01 & 0.98 & 0.40 & 32 & 5 & 2.49 & 0.78 & 0.96 & 0 & 37 & 5 & 4.05 & 0.54 & 0.97 & 0 \\ 
\bottomrule
\end{tabularx}
   \begin{tablenotes}
      \item{Respondents to a post-event evaluation survey indicate their level of agreement on a six-point Likert-type scale to a range of statements (letters correspond to panels in Figure \ref{fig:survey}). Results are for hack weeks held in 2016 and 2017 for AHW and in 2016 for GHW and NHW, respectively. The number of respondents for each question is included in the table above. Additional figures showing the distributions of responses for each question are available in the SM, Section 6.}
      \item[\emph{a}]{While the range of Likert items was the same for every question, the degrees of freedom for $\chi^2$ and Cram\'{e}r's V may differ depending on the number of items with non-zero responses, and the number of non-zero items in the independent variable.}
      \item[\emph{b}]{The degrees of freedom for Cram\'{e}r's V corrected according to~\citep{bergsma2013}}
      \item[\emph{c}]{We defined ``minority'' with respect to the field of research and grouped women and non gender-conforming participants into this category.}

@@

@@

Given the broad diversity in skills and backgrounds of participants admitted to our hack weeks, a dependence of learning outcomes and teaching on career stage is plausible. At the same time, we put forth the hypothesis earlier in the this paper that peer learning is a major mode of knowledge transfer during hack weeks, and that this type of learning takes place at all career stages. However, we find no strong evidence for a significant difference between early-career researchers and senior participants for any of the three hack weeks for questions regarding learning outcomes of new tools and topics (a), improvements in day-to-day research (b) or overall improvements in science (c) (all statistical results are presented in Table \ref{tab:survey}). Only for GHW we find a weak indication that early-career researchers agree more strongly that the hack week improved their day-to-day research (p=0.02) with a large effect size of $\phi_c = 0.4$ (following~\citep{cohen1988}). 
Similarly, we find no strong indication that late-career researchers self-report a higher level of teaching at hack weeks compared to early-career researchers (d).
